---
title: "piekarski_gov51_final_paper"
format: html
editor: visual
---

# Can We Predict What NHL Players Are Likely To Have A Breakout Season? Evaluating MONEYPUCK

#### By: Sage Piekarski

Gov 51: Data Analysis and Politics

Professor Cunningham

2026 - 05 - 10

## 1 Introduction

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

## 2 Data and Measurement

### 2.1 The MONEYPUCK Dataset

```{r}
#clear the workspace and clear console
rm(list=ls())
cat('\014')

#Install and load packages
if (!require(tidyverse)) install.packages("tidyverse"); library(tidyverse)
if (!require(dplyr)) install.packages("dplyr"); library(dplyr)
if (!require(glmnet)) install.packages("glmnet");library(glmnet)
if (!require(knitr)) install.packages("knitr");library(knitr)
if (!require(scales)) install.packages("scales");library(scales)

#read csv
skaters_2008_to_2024 <- read.csv("Data/Raw/skaters_2008_to_2024.csv")

#glimpse & understand data
glimpse(skaters_2008_to_2024)
```

### 2.2 Data Cleansing and Breakout Idenitifcation

You can add options to executable code like this

```{r}
#clean the data
player_year <- skaters_2008_to_2024 %>% 
  filter(situation == "all") %>% 
  arrange(playerId, season) %>% 
  group_by(playerId) %>% 
  mutate(next_goals = lead(I_F_goals),
         goal_change = next_goals - I_F_goals) %>% 
  ungroup() %>% 
  filter(!is.na(goal_change), games_played>0) %>% 
  mutate(goals_per_game = I_F_goals / games_played,
        shots_per_game = I_F_shotsOnGoal / games_played,
        xgoals_per_game = I_F_xGoals / games_played)

#define a breakout player
breakout_threshold <- quantile(player_year$goal_change, 0.90, na.rm = TRUE)
player_year <- player_year %>% 
  mutate(breakout = ifelse(goal_change >= breakout_threshold, 1, 0))

#check
mean(player_year$breakout)
breakout_threshold
```

The `echo: false` option disables the printing of code (only output is displayed).

### 2.3 Descriptive Statistics

```{r}
#summary statistics table
player_year |> 
  summarise(`N` = n(),
            `Breakout Rate (%)` = round(mean(breakout) * 100, 1),
            `Mean Goals` = round(mean(I_F_goals, na.rm = TRUE), 1),
            `Mean xGoals` = round(mean(I_F_xGoals, na.rm = TRUE), 1),
            `Mean Shots on Goal` = round(mean(I_F_shotsOnGoal, na.rm = TRUE), 1),
            `Mean Games Played` = round(mean(games_played, na.rm = TRUE), 1),
            `Share Defensemen (%)` = round(mean(position == "D")*100, 1)) |>
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") |>
  kable(caption = "Summary Statistics, MoneyPuck Dataset")

```

```{r}
#Breakout Rate By Current Goal Total
player_year |> 
  mutate(goals_section = cut(I_F_goals, breaks = c(-Inf, 5, 15, 30, Inf), labels = c("0-5", "6-15", "16-30", "30+"))) |>
  group_by(goals_section) |>
  summarise(`N` = n(),
            `Breakout Rate (%)` = round(mean(breakout) * 100, 1))|>
  kable(caption = "Breakout Rate By Current Season Goal Total")
```

```{r}
#correlations
cor_table <- player_year %>% 
  select(I_F_goals, next_goals, goal_change, I_F_xGoals, I_F_shotsOnGoal, games_played) %>%
  cor(use = "complete.obs")
cor_table
```

```{r}
#For Positon Assumption
position_summary <- player_year %>% 
  group_by(position) %>% 
  summarise(Rate = mean(breakout), N = n()) %>% 
  arrange(desc(Rate))

ggplot(position_summary, aes(x = Rate, y = reorder(position, Rate), fill = position == "D")) +
geom_col(width = 0.6) +
geom_text(aes(label = percent(Rate, accuracy = 0.1)), hjust = -0.1, size = 3.5) +
scale_fill_manual(values = c("TRUE" = "#2166AC", "FALSE" = "#B2C8D9"), guide = "none") +
scale_x_continuous(labels = percent_format(), limits = c(0, 0.65)) +
labs(
title = "Breakout Rates Vary Substantially by Position",
subtitle = "Share with top 10% goal increase next season, unadjusted",
x = "Breakout rate",
y = NULL,
caption = "Source: MoneyPuck"
) +
theme_minimal(base_size = 11) +
theme(panel.grid.minor = element_blank())

```

```{r}
#create a figure
ggplot(player_year, aes(x = I_F_xGoals, y = goal_change)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Current Season xGoals vs Next Season Goal Increase",
       x = "Current Season Expected Goals",
       y = "Next Season Goal Change")+
  theme_minimal()
```

```{r}
#better figure
ggplot(player_year, aes(x = xgoals_per_game, y = goal_change)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "xGoals per Game vs Next Season Goal Increase",
       x = "xGoals per Game",
       y = "Next Season Goal Change")+
  theme_minimal()
```

## 3 Prediction: From OLS to LOGIT to LASSO

### 3.1 Train/Test Split

3 Prediction:

```{r}
#trail/test split
set.seed(123)
train_info <- sample(1:nrow(player_year), 0.7*nrow(player_year))
train <- player_year[train_info, ]
test <- player_year[-train_info, ]
cat("Training set:", nrow(train), "observations\n")
cat("Test set:", nrow(test), "observations\n")


```

### 3.2 Baseline: OLS (Linear Probability Model)

```{r}
#basic ols
mod1 <- lm(breakout ~ I_F_xGoals + I_F_shotsOnGoal + games_played + position, 
           data = train)
summary(mod1)

# Predict on test set
test$ols_pred <- predict(mod1, newdata = test)
test$ols_class <- ifelse(test$ols_pred > 0.10, 1, 0)

# Performance
ols_rmse <- sqrt(mean((test$breakout - test$ols_pred)^2))
ols_accuracy <- mean(test$ols_class == test$breakout)

cat(sprintf("OLS — Test RMSE: %.4f | Accuracy: %.3f\n", ols_rmse, ols_accuracy))
table(predicted = test$ols_class, actual = test$breakout)
```

### 3.3 LOGIT

```{r}
#logit model because binary
logit <- glm(breakout ~ xgoals_per_game + shots_per_game + games_played + position,
             data = train,
             family = "binomial")
summary(logit)

#OOS pred

test$pred_prob <- predict(logit, newdata = test, type = "response")
test$pred_class <- ifelse(test$pred_prob > 0.10, 1, 0)

logit_rmse <- sqrt(mean((test$breakout - test$pred_prob)^2))
logit_accuracy <- mean(test$pred_class == test$breakout)

cat(sprintf("LOGIT — Test RMSE: %.4f | Accuracy: %.3f\n", logit_rmse, logit_accuracy))
table(predicted = test$pred_class, actual = test$breakout)
```

### 3.4 LASSO

```{r}
#LASSO
x_train <- model.matrix(breakout ~ xgoals_per_game + shots_per_game + games_played + position, 
                        data = train)
y_train <- train$breakout

x_test <- model.matrix(breakout ~ xgoals_per_game + shots_per_game + games_played + position, 
                       data = test)

#CV LASSO for lambda
set.seed(42)
lasso_cv <- cv.glmnet(x_train, y_train, family = "binomial", alpha =1)
plot(lasso_cv)

lambda1 <- lasso_cv$lambda.min

#ideal lambda incl
lasso1 <- glmnet(x_train, y_train, family = "binomial", alpha =1, lambda = lambda1)
coef(lasso1)

#predict with lambda min
test$lasso_prob <- as.vector(predict(lasso1, newx = x_test, type = "response"))
test$lasso_class <- ifelse(test$lasso_prob > 0.10, 1, 0)

lasso_rmse <- sqrt(mean((test$breakout - test$lasso_prob)^2))
lasso_accuracy <- mean(test$lasso_class == test$breakout)

cat(sprintf("LASSO — lambda.min: %.4f | Test RMSE: %.4f | Accuracy: %.3f\n", 
            lambda1, lasso_rmse, lasso_accuracy))

table(predicted = test$lasso_class, actual = test$breakout)

# Which variables survive?
lasso_coef <- coef(lasso_cv, s = "lambda.min")
nonzero <- lasso_coef[lasso_coef[, 1] != 0, , drop = FALSE]
cat(sprintf("Nonzero coefficients: %d of %d\n"
, nrow(nonzero) - 1, ncol(x_train)))
```

### 3.5 Comparing the Models

```{r}
tibble(Model= c("OLS (no regularization)", "Logit", "LASSO (logit, lambda.min)"),
       `Test RMSE` = round(c(ols_rmse, logit_rmse, lasso_rmse), 4),
       `Test Accuracy`= percent(c(ols_accuracy, logit_accuracy, lasso_accuracy), accuracy = 0.1)) %>%
  kable(caption = "Out-of-sample prediction performance: OLS vs. Logit vs. LASSO")
  
```

## 4 Fairness: Does the Modell Err Equally?

```{r}
# Use LASSO predictions (lambda.min)
test |>
  group_by(position) |>
  summarise(N = n(),
            `True positive rate` = round(mean(pred_class[breakout == 1] == 1), 3),
            `False positive rate`= round(mean(pred_class[breakout == 0] == 1), 3),
            `Accuracy`= round(mean(pred_class == breakout), 3),
            .groups = "drop")|>
kable(caption = "Logit prediction error rates by position False positive rate = share predicted breakout who did not breakout.")
```

## 5 Top Predicted Breakouts

```{r}
#get players
predicted_breakouts <- test %>% 
  arrange(desc(pred_prob)) %>% 
  select(name, season, position, pred_prob, breakout) %>% 
  head(25)
knitr::kable(predicted_breakouts)
```

## 6 Limitations

## 7 Conclusion
